# The Basics of R {#basics}

The MOST IMPORTANT thing about writing code is to comment it with descriptions
of what each portion of code does, not only for your own sanity when
revisiting code, but also to allow others who may use your code to make sense
of it. R will ignore lines that start with a pound sign/"hashtag" as comments

```{r}
# R won't run this line
"But R will run this line, and each uncommented one after this, in order within your file."
```

You can write code and run it directly from within R or RStudio (Ctrl+Enter),
or you can run code from R scripts by using the function `source`

```{r, eval=FALSE}
source("welcome.R")
```

But this won't work until we are in the same directory (folder) as the file...

First, let's find out where we are. We do this with the `getwd` function. As a brief aside: in R, you need to follow most functions with parentheses,
although sometimes you won't need to put anything in these parentheses
```{r}
getwd()
```

I forgot- I put the welcome script within the folder "~/day1/Welcome/", `setwd` there. If you're using RStudio, as you can autocomplete filenames with the tab key.
```{r, eval=FALSE}
setwd("bootcampR/day01/Welcome/")
```

If you're not a fan of doing things through the command line/script (soon you
will be), there are GUI ways to do this, too: `choose.dir` (in Windows)
```{r, eval=FALSE}
setwd(choose.dir())
```

Now we can finally run the welcome script
```{r}
source("../../day01/Welcome/welcome.R")
```

Just like you can use a GUI to choose a folder, you can do the same for files
```{r, eval=FALSE}
source(choose.files())
```

Alright, now that we're done with that, let's go back to the parent directory
with the `../` shortcut to say "go up one directory in the file structure"
```{r, eval=FALSE}
setwd("../")
```

## Math

At it's heart, R is just a big calculator...

Addition
```{r}
1 + 2
```

Subtraction
```{r}
2 - 1
```

Multiplication
```{r}
2 * 3
```

Division
```{r}
4 / 2
```

Exponents
```{r}
3 ^ 2
```

Modulo (a.k.a. Remainder)
```{r}
5 %% 2
```

Integer (Floor) Division
```{r}
5 %/% 2
```

Math in R follows the same basic order of operations that you mathematicians 
follow in your everyday work:

1. Exponents (right to left)
2. Modulo (left to right)
3. Multiplication & Division (left to right)
4. Addition & Subtraction (left to right)

```{r}
8 + 16 * 2 / 4 ^ 2 %% 3
```

You can, however, alter the order of operations with parentheses in creative 
places:
```{r}
((8 + 16) * 2 / 4) ^ (2 %% 3)
```

R also has lots of built-in math functions (more on functions later)
```{r}

abs(-3)       # absolute value
sqrt(64)      # square root
factorial(4)  # factorial
choose(8, 2)  # n choose m
```

The `log` function computes natural logs by default, but you can also select
other options for this using the `base = ...` argument in the function, or you 
can compute base-10 log using the function `log10`
```{r}
log(30)            # natural log (base e)
log(64, base = 8)  # log, base 8
log10(100)         # log, base 10
```

$\pi$ is a pre-defined constant in R, accessible by just typing `pi`
```{r}
pi
```

Trigonometric functions use radians by default
```{r}
sin(pi / 2)
cos(pi)
tan(pi / 4)
```

## Variables

Variables are an easy way to store pretty much anything that you may want to
use later in your code. Like in Python, you can use the equal sign for this
```{r}
x = 42
```

You may have noticed by now that I use spaces between pretty much everything.
This isn't necessary, but it is stylistically preferred. For instance, this
will do the same thing as the above,  lineButWhenYouHaveNoSpacesBetweenYourOperationsThingsCanGetHardToRead.
```{r}
x=42
```

However, the preferred notation for variable assignment in R is `<-`
This is because, by default, the item on the right is always assigned to
the item on the left. The arrow allows you to keep track of this better
```{r}
y <- 10
```

You can see what's stored in the variable using `print`
```{r}
print(y)
```

But, unlike in Python, you don't have to actully type `print`; this is
actually what you've been doing all along with the other examples so far
```{r}
y
```

And, if you'd like to assign and print your variable in the same line, use 
parentheses:
```{r}
(z <- -2)
```

When you have a variable, you can do anything with it that you could with the
original value
```{r}
z + 10
x + y
```

And we can assign the results of operations directly to new variables
(you can name variables anything you want, with some exceptions, including:
no spaces, no quotation marks, no operators like +-/*^%><=!)
```{r}
My_Variable <- 10 + y
```

And you can also re-assign values to variables- for instance, x was previously
set to 42; let's see what happens when we assign 12 to it...
```{r}
x
x <- 12
x
```

Values of variables are calculated at assignment, and are (almost never)
recalculated on the fly (if a variable was used to assign a value to another
variable, changing the value of one after assignment will not affect the value
of the other- they are immutable and not passed by reference for assignment)
```{r}
(NewVar <- x + y)
y <- 24
NewVar
```

Variables can be of many different types, or classes, in R:
Null (empty):
```{r}
i <- NULL
class(i)
```

Numeric (float / double):
```{r}
j <- 2
class(j)
```

Integer (must be stated strictly):
```{r}
k <- as.integer(4)
class(k)
```

Logical (True or False, a.k.a. Boolean):
```{r}
l <- TRUE
class(l)
```

Character (string):
```{r}
m <- "a"
class(m)
```

## Functions

If you have certain tasks that you do a lot in R, the best way to save time is
to automate these tasks by writing functions. This may sound daunting, but it
really isn't- functions have only three key parts: argument, body, and return.

The way you create a function in R is very similar to how you would create a
variable. For this function, we are just multiplying a number by two:
```{r}
multiply_by_two <- function(argument) {
  x <- argument * 2
  return(x)
}

multiply_by_two(3)
```

You can also make functions that take multiple inputs, or arguments (but you
don't have to name them "argument" like I did in the first example; you do
have to use the word `function` when creating a function though, and you
should also use `return` at the end of the function).
```{r}
multiply <- function(a, b) {
  return(a * b)
}

multiply(3, 4)
```

But it is important to note that named arguments are only accessible within
the function where they appear
```{r, eval=FALSE}
a
b
```

...the same goes for variables that are initialized within functions
```{r, eval=FALSE}
my_function <- function(c) {
  d <- c ^ c
  return(d)
}

my_function(3)

d
```

This is due to a phenomenon called lexical scoping: variables will only be
available or changeable in the the context in which they are created (and all
contexts that are built from that context). This means that a variable that is
initialized outside of a function (in the "global" scope) will be available
in both the global and function scope; remember that we initialized z to -2
```{r}
my_other_function <- function(arg) {
  return(arg * z)
}

my_other_function(4)
```

...but if we change the value of a global variable in the function scope, this
will not carry back over to the more general global scope
```{r}
my_other_other_function <- function(f) {
  z <- f ^ 3
  return(z)
}

my_other_other_function(2)

z
```

One last thing about functions for today: you can save the return values of
functions as variables- this will make it so that you don't have to
recalculate functions each time you want to use the same output
```{r}
h <- choose(20, 2)
h
```

There's a lot more that you can do with functions, and we'll come back to
functions throughout the course as we get deeper into R's capabilities!

* It's best practice with R code to always end your script with an empty line
